\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization

### 2010-04-20 honors_thesis/ppsn_paper/paper.tex
Got rid of figure in intro, almost down to 10 pages

These previous parallel techniques distribute the computation needed by the
particles in the swarm over the available processors.  If more processors are
available, these techniques increase the number of particles in the swarm,
either by adding individual particles or by adding entire new sub-swarms, but
in all cases the number of processors does not exceed the number of particles.
The number of iterations of PSO that the algorithms can perform is thus
inherently limited by the time it takes to evaluate the objective
function---additional processors add more particles, but do not make the
iterations go any faster.

For many functions there comes a point of diminishing returns with respect to
adding particles.  In \fig{evals-sphere} we show the value obtained after
50,000 function evaluations (not iterations) as a function of swarm size for
the well-know benchmark function Sphere (20 dimensions, reporting the average
of twenty runs).  Increasing the swarm size from 5 to 10 has a significant
effect on the value obtained.  However, increasing the swarm size from 16 to 30
makes the algorithm less efficient, that is, it reduces the progress the
algorithm makes per evaluation.  Other functions show similar trends, though
often the optimal swarm size is larger.  For this reason, previous
work has recommended the use of a swarm size of 50 for
PSO~\cite{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least some
cases, adding particles indefinitely will not yield an efficient
implementation.

\begin{figure}
  \centering
  \includegraphics[width=.45\columnwidth]{evals_sphere}
  \caption{Function Sphere with various swarm sizes. Error bars show median and
  10th and 90th percentiles.}
  \label{fig:evals-sphere}
\end{figure}

@@

To solve this problem, we propose a method for performing two iterations of PSO
at the same time in parallel that we call speculative evaluation.  The name
comes from an analogy to speculative execution (also known as branch
prediction) a technique commonly used in processors.  Modern processors, when
faced with a branch on which they must wait (e.g., a memory cache miss), guess
which way the branch will go and start executing, ensuring that any changes
can be undone.  If the processor guesses right, execution is
much farther ahead than if it had idly waited on the memory reference.  If
it guesses wrong, execution restarts where it would have been anyway.

In this paper we show that the results of standard PSO can be reproduced
\emph{exactly}, two iterations at a time, using a speculative approach similar
to speculative execution. We prove that the standard PSO equation can be
factored such that a set of speculative positions can be found which will
\emph{always} include the position computed in the next iteration.  By
computing the value of the objective function for each of the speculative
positions at the same time the algorithm evaluates the objective function for
the current position, it is possible to know the objective function values for
both the current and the next iteration at the same time.
The resulting implementation
runs efficiently on large clusters where the number of processors is much
larger than a typical or reasonable number of particles, producing better
results in less ``wall-clock'' time.

The balance of this paper is organized as follows. \sec{pso} describes the
particle swarm optimization algorithm.  \sec{sepso} describes how speculative
evaluation can be done in parallel PSO to perform two iterations at once, then
presents a comparison of the speculative algorithm with the original PSO.  In
\sec{conclusion} we conclude.

@@

The PSO algorithm can be trivially parallelized by distributing the computation
needed for each particle across processors.  But
for some functions, as shown in \fig{evals-sphere},
adding particles yields diminishing returns.
That is, adding processors does not help reach any given level of fitness
appreciably faster.  Instead of adding particles, speculative
evaluation performs iterations two at a time.

@@

To see the value of this refactoring, suppose that $800$ processors are
available, and that the evaluation of the function to be optimized takes one
hour.  If we only want a swarm of $100$ particles, $700$ of the processors
would be sitting idle for an hour at every iteration, and it would take two
hours to run two iterations.  If instead we perform speculative evaluation,
sending each of the $\ofunc(\specpos_{t+1}^{c})$ to be computed at the same
time as $\ofunc(\ppos_{t})$, we could create a swarm of size 100, each particle
with $7$ speculative evaluations ($700$ processors dedicated to speculative
evaluation), thus using all $800$ processors and performing two iterations in
one hour.

>>>>

These previous parallel techniques distribute the computation needed by the
particles in the swarm over the available processors.  If more processors are
available, these techniques increase the number of particles in the swarm.  The
number of iterations of PSO that the algorithms can perform is thus inherently
limited by the time it takes to evaluate the objective function---additional
processors add more particles, but do not make the iterations go any faster.

For many functions there comes a point of diminishing returns with respect to
adding particles.  Very small swarms do not produce enough exploration to
consistently find good values, while large swarms result in more exploration
than is necessary and waste computation.  For this reason, previous work has
recommended the use of a swarm size of 50 for
PSO~\cite{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least some
cases, adding particles indefinitely will not yield an efficient
implementation.

@@

To solve this problem, we propose a method for performing two iterations of PSO
at the same time in parallel that we call speculative evaluation.  The name
comes from an analogy to speculative execution (also known as branch
prediction), a technique commonly used in processors.  Modern processors, when
faced with a branch on which they must wait (e.g., a memory cache miss), guess
which way the branch will go and start executing, ensuring that any changes can
be undone.  If the processor guesses right, execution is much farther ahead
than if it had idly waited on the memory reference.  If it guesses wrong,
execution restarts where it would have been anyway.

We show that the results of standard PSO can be reproduced \emph{exactly}, two
iterations at a time, using a speculative approach similar to speculative
execution. We prove that the standard PSO equation can be factored such that a
set of speculative positions can be found which will \emph{always} include the
position computed in the next iteration.  By computing the value of the
objective function for each of the speculative positions at the same time the
algorithm evaluates the objective function for the current position, it is
possible to know the objective function values for both the current and the
next iteration at the same time.  The resulting implementation runs efficiently
on large clusters where the number of processors is much larger than a typical
or reasonable number of particles, producing better results in less
``wall-clock'' time.

The balance of this paper is organized as follows. \sec{pso} describes the
particle swarm optimization algorithm.  \sec{sepso} describes how speculative
evaluation can be done in parallel PSO to perform two iterations at once.
\sec{extra} then introduces a variant of the speculative algorithm that in some
cases performs better.  In \sec{results} and \sec{conclusion} we present our
results and conclude.

@@

The PSO algorithm can be trivially parallelized by distributing the computation
needed for each particle across processors.  But for some functions adding
particles yields diminishing returns.  That is, adding processors does not help
reach any given level of fitness appreciably faster.  Instead of adding
particles, speculative evaluation performs iterations two at a time.

@@

To see the value of this refactoring, suppose that $800$ processors are
available, and that the evaluation of the objective function takes one hour.
If we only want a swarm of $100$ particles, $700$ of the processors would be
sitting idle for an hour at every iteration, and it would take two hours to run
two iterations.  If instead we perform speculative evaluation, sending each of
the $\ofunc(\specpos_{t+1}^{c})$ to be computed at the same time as
$\ofunc(\ppos_{t})$, we could create a swarm of size 100, each particle with
$7$ speculative evaluations ($700$ processors dedicated to speculative
evaluation), thus using all $800$ processors and performing two iterations in
one hour.
